Unlocking Sensitivity for Visibility-based Estimators of the 21\,cm Reionization Power Spectrum

### 2017-11-28 Rotation paper/revision_1.tex
save

Visibility-based pipelines are not fundamentally limited to using only fully redundant baselines. In fact, most sensitivity forecasts to date do include partial redundancy \citep{Pobersens, HERA, JoshAntPos}. \textbf{A number of methods to produce power spectrum measurements using partially redundant baselines has been studied in recent years. A power-spectrum estimator based on visibilities gridded in the $uv$-plane was introduced in \cite{Choudhury2014Estimators, Choudhury2016Taperingb, Choudhury2016Taperinga}, and compared to a simple pair-baseline estimator in \cite{Choudhury2014Estimators}. Due to considerations of computational complexity, the authors favored grid-based estimator over pair-baseline estimators despite the higher accuracy of the latter.  \cite{wterm} proposed another approach based on gridded visibilities in sky-tracking measurements. 
%Arrays capable of tracking include MWA and LOFAR, but not drift-scan telescopes such as PAPER, HERA, and potentially SKA-low.
\cite{CTrott} studied three observing strategies (tracking, drift scan, and a combination of the two) and developed a pair-visibility based approach to power spectrum estimation.
%and concluded that zenith drift scan was the preferred observing strategy.
In this paper, we extend the above works and introduce a non-gridded \emph{baseline}-pair power spectrum estimator that can be applied to both tracking-capable arrays and drift scan-only arrays. Our formalism also provides a way to identify which baseline pairs to cross-correlate, overcoming the aforementioned computational disadvantage of baseline-based estimators relative to grid-based estimators.
Furthermore, compared to
%the pair-baseline estimators studied in
\cite{Choudhury2014Estimators} and \cite{CTrott}, our analysis 
more explicitly treats the curved nature of the sky as well as its spectral dimension.
%of the sensitivities indicate the importance of the three dimensional nature of the sky, resulting in us directly addressing curved sky effects in our formalism. 
}

@@

The dashed lines represent the values when only the equivalent baseline-pairs are used. The small and unfractured HERA-37, with no gap (like in HERA-350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$. \textbf{As seen in the scatter plot of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$ in Fig. \ref{fig:pairplot}, these are also the shorter baselines, which tend to contribute sensitivity to lower $k$-modes.} The fractured HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. \textbf{However, compared to PAPER, HERA's boost in sensitivity is mostly attributable to the longer baselines---and thus high $k$---modes (see Fig. \ref{fig:pairplot}). } The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``unfractured'' HERA-37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}

>>>>

Visibility-based pipelines are not fundamentally limited to using only fully redundant baselines. In fact, most sensitivity forecasts to date do include partial redundancy \citep{Pobersens, HERA, JoshAntPos}. \textbf{A number of methods to produce power spectrum measurements using partially redundant baselines has been studied in recent years. A power-spectrum estimator based on visibilities gridded in the $uv$-plane was introduced in \cite{Choudhury2014Estimators, Choudhury2016Taperingb, Choudhury2016Taperinga}, and compared to a simple pair-baseline estimator in \cite{Choudhury2014Estimators}. Due to considerations of computational complexity, the authors favored grid-based estimator over pair-baseline estimators despite the higher accuracy of the latter.  \cite{wterm} proposed another approach based on gridded visibilities in sky-tracking measurements. 
%Arrays capable of tracking include MWA and LOFAR, but not drift-scan telescopes such as PAPER, HERA, and potentially SKA-low.
\cite{CTrott} studied three observing strategies (tracking, drift scan, and a combination of the two) and developed a pair-visibility based approach to coherently combine visibilities gridded on the $uv$-plane for power spectrum estimation.
%and concluded that zenith drift scan was the preferred observing strategy.
In this paper, we extend the above works and introduce a non-gridded \emph{baseline}-pair power spectrum estimator that can be applied to both tracking-capable arrays and drift scan-only arrays. Our formalism also provides a way to identify which baseline pairs to cross-correlate, overcoming the aforementioned computational disadvantage of baseline-based estimators relative to grid-based estimators.
Furthermore, compared to
%the pair-baseline estimators studied in
\cite{Choudhury2014Estimators} and \cite{CTrott}, our analysis 
more explicitly treats the curved nature of the sky as well as its spectral dimension.
%of the sensitivities indicate the importance of the three dimensional nature of the sky, resulting in us directly addressing curved sky effects in our formalism. 
}

@@

The dashed lines represent the values when only the equivalent baseline-pairs are used. The small and unfractured HERA-37, with no gap (like in HERA-350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$. \textbf{As seen in the scatter plot of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$ in Fig. \ref{fig:pairplot}, these are also the shorter baselines, which tend to contribute sensitivity to lower $k$-modes.} The fractured HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. \textbf{However, compared to PAPER, HERA's boost in sensitivity is mostly attributable to the longer baselines---and thus high $k$-modes (see Fig. \ref{fig:pairplot}). } The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``unfractured'' HERA-37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}
